---
title: "Exam 2"
output: html_document
---

### Econ B2000, MA Econometrics
### Kevin R Foster, the Colin Powell School at the City College of New York, CUNY
### Fall 2024

*The questions are worth 120 points.  You have 120 minutes to do the exam, one point per minute.*
*All answers should submitted electronically. Please submit all relevant computer files as a Slack message to me (perhaps just a message with GitHub link). I prefer .Rmd files __along with knit output,__ md or html is fine. Graphs should be 'export'. Or you can save output and code as pdf or rtf if you must.*
*You may refer to your books, notes, calculator, computer, etc.  The exam is "open book."*
*However, you must not refer to anyone else, either in person or electronically, during the exam time. For instance, since these exam questions are newly created, posting questions or copying answers on Discord or WhatsApp groups or online homework helping sites or forums (such as Chegg, Yahoo answers or others) is a violation. Don't upload to public GitHub site until end of exam.*
*You must do all work on your own.  Cheating is harshly penalized.*
*Good luck.  Stay cool.*

1. (10 points) This does not require any work in R (although you might find it convenient, I'm not stopping you). It uses a particular subset of the data that you can't replicate, you should be able to answer the question with just the information provided. I estimated a simple regression where people's reported mental health, Y, was a function of their age, including squared and cubed terms. Y is the number of days, of last 30, where person's mental health was poor.

$$ Y = \beta_{0} + \beta_{1} Age + \beta_{2} Age^{2} + \beta_{3} Age^{3} + \epsilon  $$
The estimated coefficients are 
$$ \hat{\beta_{0}} = 12.18 $$
$$ \hat{\beta_{1}} = -0.308 $$ 
$$ \hat{\beta_{2}} = 0.0048 $$ 
$$ \hat{\beta_{3}} = -0.00003 $$ 

What is the prediction for a person who is 25 years old? For a person who is 45 years old?
- The prediction for a person who is 25 years old is 7.01 as when you substitue the corresponding values for beta and age, you get: Y= 12.18-.308(25)+.0048(25^2)-.00003(25^3) which equates to 7.01. The prediction for a person who is 45 years old is 5.31 because once again, when substituted, we get Y= 12.18-.308(45)+.0048(45^2)-.00003(45^3) which results in 5.31.


### Background
I provide a small version of the BRFSS data. We'll look at some of the determinants of mental health. We'll focus on the variable, `MentalHealth`, which answers the question, "Now thinking about your mental health, which includes stress, depression, and problems with emotions, for how many days during the past 30 days was your mental health not good?"

2. (15 points) First with the continuous version of the measure, would like to understand the relative importance of income and education in predicting mental health. Run a regression of the form, `ols_model1 <- lm(MentalHealth ~ AgeBracket + Educ + IncomeBracket)` and perform a hypothesis test for the joint hypothesis that all of the coefficients on education are zero. What is the p-value? What about another joint hypothesis that all the coefficients on income are zero? Explain.

```{r}
library(car)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(haven)
library(lmtest)  
library(vip) 

load("/Users/ari3/Downloads/brfss_for_exam2_fall24 (1).RData")
ls()

# Model 1: Basic regression
ols_out1 <- lm(MentalHealth ~ AgeBracket + Educ + IncomeBracket, data = brfss_sm)
summary(ols_out1)
coef(ols_out1)


# Load necessary library
library(car)

# Joint hypothesis test for education coefficients
test_education <- linearHypothesis(
  ols_out1,
  c(
    "Grades 1 through 8 (Elementary)` = 0",
    "Grades 9 through 11 (Some high school)` = 0",
    "Grade 12 or GED (High school graduate)` = 0",
    "College 1 year to 3 years (Some college or technical school)` = 0",
    "College 4 years or more (College graduate)` = 0"
  )
)

# Print the results
print(test_education)


## Education is 0
test_education <- linearHypothesis(ols_model, 
                                   c("Grades 1 through 8 (Elementary) = 0",
                                     "Grades 9 through 11 (Some high school) = 0",
                                     "Grade 12 or GED (High school graduate) = 0",
                                     "College 1 year to 3 years (Some college or technical school) = 0",
                                     "College 4 years or more (College graduate) = 0"))

print("Joint Hypothesis Test for Education Coefficients:")
print(test_education)
## When all the coefficients for education is a zero


## Income is 0
test_income <- linearHypothesis(ols_model, 
                                 c("HH income 10k - 15k = 0",
                                   "HH income 15k - 20k = 0",
                                   "HH income 20k - 25k = 0"
                                   "HH income 25k - 35k = 0"
                                   "HH income 35k - 50k = 0"
                                   "HH income 50k - 75k = 0"
                                   "HH income 75k - 100k = 0"
                                   "HH income 100k - 150k = 0"
                                   "HH income 150k - 200k = 0"))

print("Joint Hypothesis Test for Income Coefficients:")
print(test_income)
```


3. (10 points) Select a subset of the data that you would like to analyze. Explain your choice and provide some summary statistics of interesting variables.

```{r}
subset_data <- brfss_sm %>% 
  filter(
    (MaritalStatus == "Married") &                          # Married people
    (Race == "Black or African American") &           # Black or African American race
    (Educ == "College 4 years or more (College graduate)")  # College 4 or more years
  )

# Check the first rows of the filtered dataset to verify
head(subset_data)
## I drilled down my data down to indidviduals who are of Black or African American race, married, and have a college education of 4 or more years. I chose this because I wanted to see how this would play a role in the change in income. I noticed majority of them have no children and live in Central City. They also all have emotional support.
```


I created a 0/1 version of Mental Health, `MentalHealth_01 <- (MentalHealth > 3)` which takes a value of 1 if a person reports 4 or more days of the past month with poor mental health, zero otherwise.

4. (25 points) Estimate a OLS model with the y-variable as `MentalHealth_01`. Perhaps include some interaction effects.
  a. Explain what variables you choose to use as predictors. Do they seem exogenous? Explain the interactions.
  b. Do your estimates seem plausible? Are the estimates each statistically significant?
  c. Construct a joint test of whether a reasonable set of coefficients (such as education or income dummies) are all zero.
  d. What are the predicted probabilities for a few particular groups?
  e. How many Type I and Type II errors are made by the model?
  
```{r}
## Creating MentalHealth_01
brfss_sm$MentalHealth_01 <- ifelse(brfss_sm$MentalHealth > 3, 1, 0)

head(brfss_sm$MentalHealth_01)

summary(brfss_sm$MentalHealth)

unique(brfss_sm$MentalHealth)

table(brfss_sm$MentalHealth <= 3)

brfss_sm$MentalHealth_01 <- ifelse(brfss_sm$MentalHealth > 3, 1, 0)

table(brfss_sm$MentalHealth_01)


##Estimating OLS Model
ols_model <- lm(MentalHealth_01 ~ AgeBracket * Educ + Race + IncomeBracket, data = brfss_sm)
summary(ols_model)

# Load necessary library
library(ggplot2)

# Example dataset (replace this with your actual data)
mental_health_counts <- data.frame(
  MentalHealth_01 = c("Good Mental Health (0)", "Poor Mental Health (1)"),
  Count = c(150, 50) # Replace with actual counts
)

# Create the bar chart
ggplot(mental_health_counts, aes(x = MentalHealth_01, y = Count, fill = MentalHealth_01)) +
  geom_bar(stat = "identity", width = 0.5) +
  scale_fill_manual(values = c("skyblue", "orange")) +
  labs(
    title = "Distribution of MentalHealth_01",
    x = "Mental Health Status",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    legend.position = "none"
  )





##A) The variables I chose were age, education levels, income, and race. I chose age as thisi plays a role in mental health since younger and older individuals face issues in this area differently. I chose Education because this also influences mental health due to its relationship with socioeconomic status. I chose race because this involves systemic and cultural impact on mental health outcomes. Lastly I chose income because impacts economic stability which is strongly linked to mental health. Age and race are exogenous because they won't change with mental health changes. There's interactions between age and education because things like higher education might have stronger effect on mental health for younger adults in comparison to older adults.

##B) It does not seem like my estimates are plausible as there are not really negatove coefficients correlated to higher education nor is there positive coefficients for higher incomce levels. It seems like these factors are not influencing mental health as they way I estimated.

##C)
# Load car package for hypothesis testing
library(car)

# Perform joint hypothesis test for education dummies
test_education <- linearHypothesis(
  ols_model,
  c(
    "`EducGrades 1 through 8 (Elementary)` = 0",
    "`EducGrades 9 through 11 (Some high school)` = 0",
    "`EducGrade 12 or GED (High school graduate)` = 0",
    "`EducCollege 1 year to 3 years (Some college or technical school)` = 0",
    "`EducCollege 4 years or more (College graduate)` = 0"
  )
)

# Print test results
print(test_education)
## The p value is less than .05 which shows that we are rejecting the null hypothesis.Therefore, it seems that education levels do have an impact on my mentalhealth_01 data.

##D)
# Example data for predictions
new_data <- data.frame(
  AgeBracket = c("25 to 29", "30 to 34"),
  Educ = c("College 4 years or more (College graduate)", "EducGrade 12 or GED (High school graduate)"),
  Race = c("Black or African American", "White"),
  IncomeBracket = c("Low", "High")
)

# Predicted probabilities
predicted_probs <- predict(ols_model, newdata = new_data, type = "response")
predicted_probs


##E)
# Predicted values (threshold of 0.5)
predicted <- ifelse(predict(ols_model, type = "response") > 0.5, 1, 0)

# Actual values
actual <- brfss_sm$MentalHealth_01

# Confusion matrix
table(Predicted = predicted, Actual = actual)

# Type I and Type II errors
type1_error <- sum(predicted == 1 & actual == 0)
type2_error <- sum(predicted == 0 & actual == 1)

cat("Type I Errors:", type1_error, "\n")
cat("Type II Errors:", type2_error, "\n")

```


5. (30 points) Estimate a  logit model, for the outcome variable `MentalHealth_01`, within your subsample.
  a. Explain what variables you choose to use as predictors. Do they seem exogenous? 
  b. Do your estimates seem plausible? Are the estimates each statistically significant?
  c. Construct a joint test of whether a reasonable set of coefficients are all zero.
  d. What are the predicted probabilities for a few particular groups?
  e. How many Type I and Type II errors are made by the model?
  f. How does this logit compare with the OLS previous?
  
```{r}
subset_data <- brfss_sm %>% 
  filter(
    (MaritalStatus == "Married") &                          # Married people
    (Race == "Black or African American") &           # Black or African American race
    (Educ == "College 4 years or more (College graduate)")  # College 4 or more years
  )

# Check the first rows of the filtered dataset to verify
head(subset_data)

# Fit the logistic regression model
logit_model <- glm(MentalHealth_01 ~ AgeBracket * Educ + Race + IncomeBracket, 
                   data = subset_data, 
                   family = binomial)

# Summarize the results
summary(logit_model)

# Load necessary library
library(ggplot2)

# Create a bar chart for the distribution of MentalHealth_01
ggplot(brfss_sm, aes(x = factor(MentalHealth_01, labels = c("Good Mental Health (0)", "Poor Mental Health (1)")))) +
  geom_bar(fill = c("skyblue", "orange")) +
  labs(
    title = "Distribution of MentalHealth_01",
    x = "Mental Health Status",
    y = "Count"
  ) +
  theme_minimal()



##A) I used similar variables; age, income, race, and education. The age and race are exogenous
##B) My estimates do seem plausible, the positive coefficients for low income and/orlow education individuals is increasing with the increase in likelihood of poor mental health. Additionally, the negative coefficients correlated to high education and high income are decreasing with the liklihood of mental health

##C) 
# Load the car package for joint hypothesis testing
library(car)

# Perform the joint test for income-related coefficients
# Load the car package
library(car)

# Perform the joint hypothesis test for income-related coefficients
test_income <- linearHypothesis(
  logit_model,
  c(
    "`HH income 15k - 20k` = 0",
    "`HH income 20k - 25k` = 0",
    "`HH income 25k - 35k` = 0",
    "`HH income 35k - 50k` = 0",
    "`HH income 50k - 75k` = 0",
    "`HH income 75k - 100k` = 0",
    "`HH income 100k - 150k` = 0",
    "`HH income 150k - 200k` = 0"
  )
)

# Print the results
print(test_income)

##D) 
new_data <- data.frame(
  AgeBracket = factor(c("25 to 29", "30 to 34"), levels = levels(subset_data$AgeBracket)),
  Educ = factor(c("College 4 years or more (College graduate)", "EducGrade 12 or GED (High school graduate)"), 
                levels = levels(subset_data$Educ)),
  Race = factor(c("Black or African American", "White"), levels = levels(subset_data$Race)),
  Income = factor(c("Low", "High"), levels = levels(subset_data$Income))
)

# Predicted probabilities
predicted_probs <- predict(logit_model, newdata = new_data, type = "response")
print(predicted_probs)
##The predicted probabilities are .25 and .52


##E)
# Predicted binary outcomes
predicted <- ifelse(predict(logit_model, type = "response") > 0.5, 1, 0)

# Actual values
actual <- subset_data$MentalHealth_01

# Confusion matrix
confusion_matrix <- table(Predicted = predicted, Actual = actual)
print(confusion_matrix)

# Calculate errors
type1_error <- sum(predicted == 1 & actual == 0)  # False positives
type2_error <- sum(predicted == 0 & actual == 1)  # False negatives

cat("Type I Errors (False Positives):", type1_error, "\n")
cat("Type II Errors (False Negatives):", type2_error, "\n")


##F) The logit model is a better choice for predicting the binary outcome variable MentalHealth_01 compared to the OLS model because it estimates probabilities that fall between 0 and 1. While the OLS model assumes a linear relationship and can sometimes produce predictions outside this range, the logit model uses a logistic function to keep predictions within realistic probability limits. This makes it easier to interpret the results when analyzing the likelihood of someone reporting poor mental health for 4 or more days in the past month. Another advantage of the logit model is that the coefficients represent changes in the log-odds of the outcome, which works better for binary data than the OLS model's coefficients, which directly measure changes in the outcome.
```

  
6. (30 points) Estimate one or more additional models with other methods (not OLS or logit) to predict mental health. Explain as in previous. Compare the different models and make judgments about strengths and weaknesses.
```{r}
# Probit model
probit_model <- glm(MentalHealth_01 ~ AgeBracket * Educ + Race + Income, 
                    data = subset_data, family = binomial(link = "probit"))

# Summary of the model
summary(probit_model)

```


